---
title: "lab_09.Rmd"
author: "Shaela, Robert, Bridget, Sydney"
date: "2023-11-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(janitor)
library(lubridate)
```

Loading all the data:
```{r}
jan_22_visitors <- read_csv("data/January_vistors.csv") |> clean_names()
feb_22_visitors <- read_csv("data/February_visitors.csv") |> clean_names()
mar_22_visitors <- read_csv("data/March_visitors.csv") |> clean_names()
apr_22_visitors <- read_csv("data/April_visitors.csv") |> clean_names()
may_22_visitors <- read_csv("data/May_visitors.csv") |> clean_names()
jun_22_visitors <- read_csv("data/June_visitors.csv")|> clean_names()
jul_22_visitors <- read_csv("data/July_visitors.csv") |> clean_names()
aug_22_visitors <- read_csv("data/August_visitors.csv") |> clean_names()
sep_22_visitors <- read_csv("data/September_visitors.csv") |> clean_names()
oct_22_visitors <- read_csv("data/October_visitors.csv") |> clean_names()
nov_22_visitors <- read_csv("data/November_visitors.csv") |> clean_names()
dec_22_visitors <- read_csv("data/December_visitors.csv") |> clean_names()
jan_23_visitors <- read_csv("data/jan_visitors_23.csv") |> clean_names()
feb_23_visitors <- read_csv("data/feb_visitors_23.csv") |> clean_names()
mar_23_visitors <- read_csv("data/mar_visitors_23.csv") |> clean_names()
apr_23_visitors <- read_csv("data/apr_visitors_23.csv") |> clean_names()
may_23_visitors <- read_csv("data/may_visitors_23.csv") |> clean_names()
jun_23_visitors <- read_csv("data/june_visitors_23.csv") |> clean_names()
jul_23_visitors <- read_csv("data/july_visitors_23.csv") |> clean_names()

```

We found three groups of inconsistencies in the header. So we're cleaning up the outliers.

#cleaning records part 1
```{r}
c_jan_22_visitors <- jan_22_visitors |> rename(first_name = namefirst, last_name = namelast, middle_initial = namemid, appointment_made_date = appt_made_date, appointment_start_date = appt_start_date, appointment_end_date = appt_end_date, appointment_cancel_date = appt_cancel_date, last_updated_by = last_updatedby, last_entry_date = lastentrydate,  visitee_last_name = visitee_namelast, visitee_first_name = visitee_namefirst, meeting_location = meeting_loc, caller_last_name = caller_name_last,  caller_first_name = caller_name_first, release_date = releasedate)
```

```{r}
c_feb_22_visitors <- feb_22_visitors |> rename(first_name = namefirst, last_name = namelast, middle_initial = namemid, appointment_made_date = appt_made_date, appointment_start_date = appt_start_date, appointment_end_date = appt_end_date, appointment_cancel_date = appt_cancel_date, last_updated_by = last_updatedby, last_entry_date = lastentrydate,  visitee_last_name = visitee_namelast, visitee_first_name = visitee_namefirst, meeting_location = meeting_loc, caller_last_name = caller_name_last,  caller_first_name = caller_name_first, release_date = releasedate)
```

```{r}
c_mar_22_visitors <- mar_22_visitors |> rename(first_name = namefirst, last_name = namelast, middle_initial = namemid, appointment_made_date = appt_made_date, appointment_start_date = appt_start_date, appointment_end_date = appt_end_date, appointment_cancel_date = appt_cancel_date, last_updated_by = last_updatedby, last_entry_date = lastentrydate,  visitee_last_name = visitee_namelast, visitee_first_name = visitee_namefirst, meeting_location = meeting_loc, caller_last_name = caller_name_last,  caller_first_name = caller_name_first, release_date = releasedate)
```

```{r}
c_apr_22_visitors <- apr_22_visitors |> rename(first_name = namefirst, last_name = namelast, middle_initial = namemid, appointment_made_date = appt_made_date, appointment_start_date = appt_start_date, appointment_end_date = appt_end_date, appointment_cancel_date = appt_cancel_date, last_updated_by = last_updatedby, last_entry_date = lastentrydate,  visitee_last_name = visitee_namelast, visitee_first_name = visitee_namefirst, meeting_location = meeting_loc, caller_last_name = caller_name_last,  caller_first_name = caller_name_first, release_date = releasedate)
```

```{r}
c_may_22_visitors <- may_22_visitors |> rename(first_name = namefirst, last_name = namelast, middle_initial = namemid, appointment_made_date = appt_made_date, appointment_start_date = appt_start_date, appointment_end_date = appt_end_date, appointment_cancel_date = appt_cancel_date, last_updated_by = last_updatedby, last_entry_date = lastentrydate,  visitee_last_name = visitee_namelast, visitee_first_name = visitee_namefirst, meeting_location = meeting_loc, caller_last_name = caller_name_last,  caller_first_name = caller_name_first, release_date = releasedate)
```

#cleaning records, part 2
```{r}
c_jun_22_visitors<- jun_22_visitors |> rename(release_date = releasedate)

c_jul_22_visitors <-jul_22_visitors |> rename(release_date = releasedate)

c_aug_22_visitors <-aug_22_visitors|> rename(release_date = releasedate)

c_sep_22_visitors <-sep_22_visitors|> rename(release_date = releasedate)

c_oct_22_visitors <-oct_22_visitors |> rename(release_date = releasedate)

c_nov_22_visitors <-nov_22_visitors |> rename(release_date = releasedate)

c_dec_22_visitors <-dec_22_visitors |> rename(release_date = releasedate)

c_jan_23_visitors <-jan_23_visitors |> rename(release_date = releasedate)

c_feb_23_visitors <-feb_23_visitors |> rename(release_date =releasedate)
```

#binding the rows of the cleaned data frames to make one big data frame with all information in it
```{r}
wh_records_combined <- bind_rows(list(c_jan_22_visitors, c_feb_22_visitors, c_mar_22_visitors, c_apr_22_visitors, c_may_22_visitors, c_jun_22_visitors, c_jul_22_visitors, c_aug_22_visitors, c_sep_22_visitors, c_oct_22_visitors, c_nov_22_visitors, c_dec_22_visitors, c_jan_23_visitors, c_feb_23_visitors, may_23_visitors, apr_23_visitors, may_23_visitors, jun_23_visitors, jul_23_visitors))

wh_records_combined
```

#creating a column for visitor name that combines first name, last name and middle initial. Creating a column for visitee name. Getting rid of unnecessary columns (x28 and x29, caller_room, release_date)
```{r}
wh_records_combined_names <- wh_records_combined |> mutate(full_name_visitor = paste(first_name, middle_initial, last_name, sep = '_'),
                              full_name_visitee = paste(visitee_first_name, visitee_last_name, sep = '_'))|>
  select(-x28, -x29,-caller_room, -release_date)
```

#cleaning the dates
```{r}
wh_combined_dates_and_names <- wh_records_combined_names|>
  mutate(c_appointment_start_date = mdy_hm(appointment_start_date), 
         c_appointment_end_date = mdy_hm(appointment_end_date),
         c_appointment_cancel_date = mdy_hm(appointment_cancel_date), 
         c_last_entry_date = mdy_hm(last_entry_date),
         c_appointment_made_date = mdy_hm(appointment_made_date), 
         c_toa = mdy_hm(toa))

```
***Milestone 2*** 
After cleaning up the data we believe we'll be able to answer question 1 about who visited the white house the most between 2022 and the first seven months of 2023. Flaws and limitations hindering us answering our questions: after cleaning up the data, we believe we will be able to answer question 1 regarding who visited the white house most in 2022 and 2023. 

We also think we'll be able to answer question 2 (who is visited most often) by observing the most frequently listed visitee. We'll also be able to answer what percentage of the visits are to the president or first lady. 

For the question regarding what time of day, day of the week, etc. is the busiest we could use the toa (time of appointment) column but there are several N/A's listed. If we find anything we will have to be cautious. Also, the date and time are combined into one column so we'll need to split them up. 

We won't be able to answer question five since we can't calculate the length of appointments. The toa is listed, but the tod (time of departure) is not. The end date for all appointments is 11:59 on the day of the appointment, but we know there's no guarantee everyone stayed until midnight for their appointment. With this, we won't be able to determine anything. 

We also might have trouble answering question six regarding how frequently does house or senate leaders visit the White House? A better question we think would be: what percentage of the logged visits are from leaders of the house or senate? We haven't fully investigated this question, but we believe we could create a list of name of high-ranked leaders (speaker of the house, minority whip, etc.) and writing code using string detect and mutate and case when to create a column that will be marked yes if the entry is a member of the house of senate leadership and no if they aren't. 

***Answering Questions***

Question 1: Who visits the most? 

Answer: Michael N Herzog visits the white house the most with 108 visits. After doing research I found out he's Israel's Ambassador to the U.S. This finding was interesting as we could explore what months he frequently visited. We won't be able to know what he came to speak with people about, but we could observe what the current climate of the country was and determine if that's something he would want to speak with representatives about. 
```{r}
wh_records_combined_names |>
  group_by(full_name_visitor) |>
  summarize(
    total_visits=n()
  ) |>
  arrange(desc(total_visits))

wh_records_combined_names
```

# Used OpenRefine to clean up some visitee names #

Question 2: Who gets the most visitors?

Answer: Ed Teleky (Director of Ceremonies at the White House), Gionelly Mills (Staff Assistant focusing on Communications, Celebrity Engagement and Events at the WH), Lydia Hecmanczuk (IT Specialist in the Executive Office of the President). The one that seemed interesting to me was Kevin Ballen, advisor to the director of the national economic council. 
```{r}
wh_records_combined_names |>
  group_by(full_name_visitee) |>
  summarise(count_visitee=n()) |>
  arrange(desc(count_visitee))
```


Question 3: How frequently does the president get visitors?

Answer: Roughly 8.9% of visitors at the WH say they are visiting just the president.
```{r}
wh_records_refined|>
  summarise(total_visits =n())

#POTUS visits(71777)/Total Number of Visits(805,945)
(71777/805945)*100
```

Question 4: What time of the day is the busiest for visitors? What day of the week? Which week of the year? (2022 = complete year)

Answer: 10 a.m. received the most number os visits. May was the month with the most visits. Week 26 was the week with the greatest amount of visits. Saturday was the day of the week with the most number of visits. 
```{r}
#adding a column for month, day of the week, week in the year, and hour in the day
wh_combined_dates_and_names_no_extra_mutated <- wh_combined_dates_and_names |>
  mutate (hour = hour(c_toa)) |>
  mutate (month = month(c_toa)) |>
  mutate (weekdays = weekdays(c_toa)) |>
  mutate (week = week(c_toa))

#most common hour- 10 A.M. (93, 720 visits)
wh_combined_dates_and_names_no_extra_mutated |>
  group_by(hour) |>
  summarise(count_visits = n())|>
  arrange(desc(count_visits))
#most common month- May (136, 318 visits)
wh_combined_dates_and_names_no_extra_mutated |>
  group_by(month) |>
  summarise(count_visits = n())|>
  arrange(desc(count_visits))
#most common week- week 19?? (Mid May)
wh_combined_dates_and_names_no_extra_mutated |>
  group_by(week) |>
  summarise(count_visits = n())|>
  arrange(desc(count_visits))
#most common weekday- Friday
wh_combined_dates_and_names_no_extra_mutated |>
  group_by(weekdays) |>
  summarise(count_visits = n())|>
  arrange(desc(count_visits))
```

Question 5: How frequently does House or Senate leadership visit the White House?

Answer: 
```{r}
wh_records_leadership <- wh_records_combined_names|>
  filter(full_name_visitor %in% c("PATTY_L_MURRAY","RICHARD_J_DURBIN","CHARLES_N_SCHUMER","ELIZABETH_N_WARREN","MARK_R_WARNER","ADDISON_M_MCCONNELL","JOHN_R_THUNE","KATHERINE_C_PHILLIPSBARRASSO","SHELLEY_M_CAPITO", "STEVE_J_SCALISE","HAKEEM_N_JEFFRIES","KATHERINE_M_CLARK","PETER_R_AGUILAR", "JAMES_E_CLYBURN","JAMES_M_JOHNSON","EMMER","STEFANIK", "PALMER"))
```

****Milestone 3***
So far, the most interesting story surrounds the category of most frequent person visited. It wouldn't be a hard news story since we haven't discovered anything weird. It would serve as a good behind-the-scenes/feature piece. 

***Final***