---
title: "lab_02"
author: "derek willis"
date: "2023-09-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

Let's keep working with the UMD courses and Maryland campaign expense data we used in the pre_lab.

```{r}
umd_courses <- read_rds("data/umd_courses.rds")
maryland_expenses <- read_csv("data/maryland_expenses.csv")
```

## Answer questions

**Q1.** How many departments have offered the courses with "Climate" in the title and at least one seat? Which department has offered the most?
**A1.** Atmospheric and Oceanic Science

```{r}
climate_courses <- umd_courses |>
  filter(str_detect(title, "Climate")) |> filter(seats >= 1) |> 
  group_by(department) |>
  summarise(
    count_classes = n()
  ) |>
  arrange(desc(count_classes))
view(climate_courses)
```

**Q2.** Of the departments listed in A2, which one is the most interesting to you, and why?
**A2.** Journalism and Economics were interesting to me mainly since I don't associate climate based things with those two departments. Although, there are some journalist who have a passion for the climate and would like to know more ways to report on it. I guess economist would need to know since the climate can impact how many is disbursed. 


**Q3** Which departments have the most courses in the Fall 2023 term? And which departments in that term have at least one syllabus posted? How are the top five departments different in each result?

**A3.** 1 (most courses) Business and Management, School of Music, English, Teaching and Learning, Policy and Leadership, and Information Studies
        2 (at least one syllabus). Business and Management, Economics, School of Music, Biological Sciences Program, and Information Studies 
        3. They are different in each result because although School of Music and English offer the most courses in the Fall, on the course website, their classes don't post their syllabi as frequently as Economics. Therefore, they drop out of the top five when we add the filter for the term AND the syllabus count. If it was only the term then they would remain in the top 5. 

```{r}
umd_courses |> filter(term == 202308) |> 
  filter(syllabus_count >= 1) |>
  group_by(department) |>
  summarise(
    count_classes = n(),
  ) |> 
  arrange (desc(count_classes))

```

**Q4.** Using the Maryland campaign expenses data, change the datatype of the expenditure_date column so that it is a date and not a chr, saving the result as the same dataframe. Then create a new column for the month based on expenditure_date. Then write code that shows total amount spent each month by Wes Moore's committee (you'll need to search the data to find the exact name). Write a sentence describing the pattern of spending over time.
**A4.** I noticed that from the end of summer to the beginning of fall they spend the most money.

```{r}
#maryland_expenses <- maryland_expenses |> mutate(expenditure_date=mdy(expenditure_date))

maryland_expenses <- maryland_expenses |> mutate(month = floor_date(expenditure_date, "month"))

maryland_expenses <- maryland_expenses |>  filter (committee_name == "Moore  Wes For Maryland") |>
 group_by (month) |>
 summarise(
  total_amount = sum(amount)
  ) 

head(maryland_expenses)
```

**Q5.** Using case_when(), create a column in the Maryland expenses data called `spending_location` indicating whether each record spent money in Maryland or outside Maryland. For records that are in Maryland, make the new column's value "In-state" and for the others, make it "Out of state". Then write code that shows the total amount of money spent in each category and describe the results. You can do this in one statement or break it up into two statements (saving the new column to your dataframe).
**A5** In-state = $77,723,146 , out-of-state = N/A

```{r}
maryland_expenses_location <- maryland_expenses |>
  mutate(
    spending_location = case_when (
        str_detect(address, " Maryland ") ~ "in-state",
        .default = "out-of-state" )
  ) |>
  group_by(spending_location) |>
  summarise(
    total_amount =sum(amount)
  )

maryland_expenses_location
```


**Q6.** Choose your own adventure: using either the UMD courses or campaign expenses data, ask and answer a question that is different than the previous ones. You can use one of the ones you provided in the pre_lab.
**A6.** What department offered the most courses in Fall 2020?
Business and Management offered the most courses in Fall 2020.
  
```{r}
umd_courses |> filter(term == 202008) |> 
  group_by(department) |>
  summarise(
    count_classes = n()
  ) |>
  arrange (desc(count_classes))

```
