---
title: "WP_Police_Homicides"
author: "Bridget Shaela Robert Sydney"
date: "2023-10-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading of the tidyverse

```{r}
library(tidyverse)
library(lubridate)
library(janitor)
```
```{r}
homicide_data <- read_csv("data/homicide-data.csv") |>
mutate(reported_date=ymd(reported_date)) |>
  mutate(reported_year = year(reported_date))
```
## Statement 1: As Baltimore has seen a stunning surge of violence, with nearly a killing each day for the past three years in a city of 600,000, homicide arrests have plummeted. (A killing each day for three straight years = 1,095 killings)

```{r}
baltimore <- homicide_data |>
  filter(city == "Baltimore")

baltimore
```
```{r}
baltimore <- homicide_data |>
 filter()

baltimore
```

##Statement 2: City police made an arrest in 41 percent of homicides in 2014; last year, the rate was just 27 percent, a 14 percentage point drop.

```{r}
###number of homicides in 2014
baltimore |>
  filter(reported_year >= "2014") |>
  filter(disposition == "closed by arrest")

```
```{r}
#counting the number of arrests made that closed open homicide cases in 2014. 
homicide_data |>
  filter(str_detect(city, "Baltimore")) |>
  filter(reported_date > "2013-12-31")|>
  filter(reported_date < "2015-01-01")|>
  filter(disposition == "Closed by arrest")|>
  summarise(
    count = n()
    )
```

```{r}
#dividing total arrests by total homicides
86/211

```
###Answer: The Washington Post correctly calculated the arrest rate for homicides in the city of Baltimore in 2014 as 40.75 or roughly 41%. Now lets calcualte the arrest rate for homicides in 2017. Was it really 27% as mentioned in the Washington Post article? 

```{r}
#counting total homicides in 2017
homicide_data |>
  filter(str_detect(city, "Baltimore")) |>
  filter(reported_date > "2016-12-31")|>
  filter(reported_date < "2018-01-01")|>
  summarise(
    count = n()
    )

```
```{r}
#counting total homicide cases closed by arrest in 2017
homicide_data |>
  filter(str_detect(city, "Baltimore")) |>
  filter(reported_date > 20161231)|>
  filter(reported_date < 20180101)|>
  filter(disposition == "Closed by arrest")|>
  summarise(
    count = n()
    )
```

```{r}
#dividing total arrests by total homicides
93/340
```
###Answer: Yes, the arrest rate for homicides in Baltimore in 2017 was 27.35 or roughly 27%. Now we want to see if homicide arrest rates plumeted over 2015 and 2015 as well. So we'll calculate those. 


```{r}
#counting total homicides in 2016
homicide_data |>
   filter(str_detect(city, "Baltimore")) |>
  filter(reported_date > 20151231)|>
  filter(reported_date < 20170101)|>
  summarise(count = n())
```
###There were 320 homicides in Baltimore in 2016. 

```{r}
#counting total homicide cases closed by arrest in 2016
homicide_data |>
  filter(str_detect(city, "Baltimore")) |>
  filter(reported_date > 20151231)|>
  filter(reported_date < 20170101)|>
  filter(disposition == "Closed by arrest")|>
  summarise(
    count = n()
  )

```

```{r}
72/320
```

###Anser: the arrest rate in 2016 was roughly 23%. That is lower than 2014's homicide arrest rate. Now we will check for 2015. 

```{r}
#counting total homicides in 2015
homicide_data |>
   filter(str_detect(city, "Baltimore")) |>
  filter(reported_date > 20141231)|>
  filter(reported_date < 20160101)|>
  summarise(count = n())
```
###There were 342 homicides repored in Baltimore in 2015

```{r}
#counting total homicide cases closed by arrest in 2015
homicide_data |>
   filter(str_detect(city, "Baltimore")) |>
  filter(reported_date > 20141231)|>
  filter(reported_date < 20160101)|>
  filter(disposition == 'Closed by arrest')|>
  summarise(count = n())
```

### There were 342 homicides in Baltimore in 2015. The police closed 87 of those cases by arresting someone. Now we need to see the arrest rate. 
```{r}
87/342

```
### Answer: The arrest rate in 2015 was 25%. So it verifies the statement 


##Statement 3: Of 50 of the nation’s largest cities, Baltimore is one of 34 where police now make homicide arrests less often than in 2014, according to a Washington Post analysis.


```{r} 
#Here we are calculating the number of homicides in major cities in 2014
homicide_2014 <- homicide_data |> 
  filter(str_detect(reported_date, "2014")) |> 
  group_by(city) |> 
  summarize(count_city = n()) |> 
  arrange(desc(count_city))

```

```{r} 
#Homicide made in 2014
homicide_2014 <- homicide_data |> 
  filter(str_detect(reported_date, "2014")) |> 
  group_by(city) |> 
  summarize(count_city = n()) |> 
  arrange(desc(count_city))

```






math:
```{r}
#Here we are finding the percent of no arrests in both closed and open cases in 2014
13+112

(125/211)*100
```
125 non arrests in 2014 out of 211 homicides, 59.24% no arrest

Data for 2017-

```{r}
#Here we are calculating the number of homicides in major cities in 2017
homicide_data |> 
  filter(str_detect(reported_date, "2017")) |> 
  group_by(city) |> 
  summarize(count_city = n()) |> 
  arrange(desc(count_city))

```


```{r}
#Here we are calculating the number of homicides in Baltimore in 2017 that were open and had no arrest
homicide_data |> 
  filter(str_detect(disposition, "Open/No arrest")) |> 
  filter(str_detect(reported_date, "2017")) |> 
  group_by(city) |> 
  summarize(count_city = n()) |> 
  arrange(desc(count_city))

```

```{r}
#Here we are calculating the number of homicides in major cities in 2017 that were closed and had no arrest
homicide_data |> 
  filter(str_detect(disposition, "Closed without arrest")) |> 
  filter(str_detect(reported_date, "2017")) |> 
  group_by(city) |> 
  summarize(count_city = n()) |> 
  arrange(desc(count_city))

```

```{r}
#Here we are finding the percent of no arrests in both closed and open cases in 2017
11+236

(247/340)*100
```

247 non arrests in 2017 out of 340 homicides, 72.65% no arrest

Answer: We can confirm that Baltimore is one of 34 major cities that makes arrest in 2017 less than 2014.