---
title: "WP_Police_Homicides"
author: "Bridget Shaela Robert Sydney"
date: "2023-10-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading of the tidyverse

```{r}
library(tidyverse)
library(lubridate)
library(janitor)
```

## Loading the WP data

```{r}
homicide_data <- read_csv("data/homicide-data.csv") |>
mutate(reported_date=ymd(reported_date)) 
```
## Statement 1: As Baltimore has seen a stunning surge of violence, with nearly a killing each day for the past three years in a city of 600,000, homicide arrests have plummeted. (A killing each day for three straight years = 1,095 killings)

```{r}
baltimore <- homicide_data |>
  filter(city == "Baltimore" & reported_date >= 2014-12-31)
  
  

baltimore
```

##Statement 2: City police made an arrest in 41 percent of homicides in 2014; last year, the rate was just 27 percent, a 14 percentage point drop.

```{r}
###calculating the number of homicides in 2014
baltimore_homicide|>
  filter(reported_date >2013-12-31)|>
  filter(reported_date <2015-01-01)|>
  summarise(count_homicides =n())

```
###Answer: there were 211 homicides in Baltimore in 2014. 
```{r}
#counting the number of arrests made that closed open homicide cases in 2014. 
baltimore_homicide|>
  filter(reported_date >2013-12-31)|>
  filter(reported_date <2015-01-01)|>
  filter(disposition == 'Closed by arrest')|>
  summarise(count_arrests = n())
```
###There were 86 homicide cases closed by arrest in 2014. 

```{r}
#dividing total arrests by total homicides
86/211

```
###Answer: The Washington Post correctly calculated the arrest rate for homicides in the city of Baltimore in 2014 as 40.75 or roughly 41%. Now lets calcualte the arrest rate for homicides in 2017. Was it really 27% as mentioned in the Washington Post article? 

```{r}
#counting total homicides in 2017
baltimore_homicide|>
  filter(reported_date >2016-12-31)|>
  filter(reported_date <2018-01-01)|>
  summarise(count_homicides = n())

```

###Answer:There were 340 homicides in Baltimore in 2017.

```{r}
#dividing total arrests by total homicides
93/340
```
###Answer: Yes, the arrest rate for homicides in Baltimore in 2017 was 27.35 or roughly 27%. Now we want to see if homicide arrest rates plumeted over 2015 and 2015 as well. So we will calculate those. 


```{r}
#counting total homicides in 2016
baltimore_homicide|>
  filter(reported_date >20151231)|>
  filter(reported_date <20170101)|>
  summarise(count_homicides = n())
```
###There were 320 homicides in Baltimore in 2016. 

```{r}
#counting total homicide cases closed by arrest in 2016
baltimore_homicide|>
  filter(reported_date >20151231)|>
  filter(reported_date <20170101)|>
  filter(disposition == 'Closed by arrest')|>

```
###There were 72 homicide cases closed by arrest in 2016

```{r}
72/320
```

###Anser: the arrest rate in 2016 was roughly 23%. That is lower than 2014's homicide arrest rate. Now we will check for 2015. 

```{r}
#counting total homicides in 2015
baltimore_homicide|>
  filter(reported_date >20141231)|>
  filter(reported_date <20160101)|>
  summarise(count_homicides = n())
```
###There were 342 homicides repored in Baltimore in 2015

```{r}
#counting total homicide cases closed by arrest in 2015
baltimore_homicide|>
  filter(reported_date >20141231)|>
  filter(reported_date <20160101)|>
  filter(disposition == 'Closed by arrest')|>
  summarise(count_arrests = n())
```

### There were 342 homicides in Baltimore in 2015. The police closed 87 of those cases by arresting someone. Now we need to see the arrest rate. 
```{r}
87/342

```
### Answer: The arrest rate in 2015 was 25%. So it verifies the statement 


##Statement 3: Of 50 of the nation’s largest cities, Baltimore is one of 34 where police now make homicide arrests less often than in 2014, according to a Washington Post analysis.

```{r}
```