---
title: "WP_Police_Homicides"
author: "Bridget Shaela Robert Sydney"
date: "2023-10-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading of the tidyverse

```{r}
library(tidyverse)
library(lubridate)
library(janitor)
library(refinr)
```

(loading the data into R and correcting the dates and making a separate year column to assist when figuring our further questions.)
```{r}
homicide_data <- read_csv("data/homicide-data.csv") |>
mutate(reported_date=ymd(reported_date)) |>
  mutate(reported_year = year(reported_date))
```
## Statement 1: As Baltimore has seen a stunning surge of violence, with nearly a killing each day for the past three years in a city of 600,000, homicide arrests have plummeted. 

(Checking to see if this statement is true, if so there should be about 1,095 killings that happened within a three year period)

Step 1: filtering to have a data set with only Baltimore in it.)
```{r}
baltimore <- homicide_data |>
  filter(city == "Baltimore")

baltimore
```
Step 2: Calculating # of homicides per year
```{r}
baltimore |>
 group_by(reported_year) |>
  summarise(
    count=n()
  ) |>
  arrange(reported_year)

baltimore
```
Answer: Total deaths over three year span (2015, 2016, 2017) was 1,002. That's close to 1,095 which we predicted would be the range of killings. 

(Checking the number of homicides divided by the number of days in the past three years.)
```{r}
1002/365*3
```

##Statement 2: City police made an arrest in 41 percent of homicides in 2014; last year, the rate was just 27 percent, a 14 percentage point drop.

Step 1: Finding homicides in 2014
Answer: 211 Homicides in Baltimore in 2014
```{r}
baltimore |>
  filter(reported_year >= '2014') |>
  group_by(reported_year) |>
  summarise(
    count_homicides=n()
  )

```

Step 2: Finding number of arrests made in homicide cases in 2014
Answer: 86
```{r}
baltimore |>
  filter(reported_year >= '2014') |>
  filter(disposition == 'Closed by arrest') |>
  group_by(reported_year) |>
  summarise (
    count_arrests=n()
  )
  
```

Step 3: Divide total arrests by total homicides
```{r}
86/211*100
```
Answer: The Washington Post correctly calculated the arrest rate for homicides in the city of Baltimore in 2014 as 40.75 or roughly 41%. 

Now lets calculate the arrest rate for homicides in 2017. Was it really 27% as mentioned in the Washington Post article? 

Step 1: Diving total arrests by total homicides for 2017. Referencing the previous data.
Answer: It's correct the arrest rate was 27.35 or roughly 27%. 

```{r}
93/340*100
```

Now we want to see if homicide arrest rates plummeted for 2015 and 2016. 

Step 1: Calculate 2015 using data from previous coding. 
Answer: 25.4%
```{r}
87/342*100
```

Step 2: Calculating 2016
Answer: 23% which is lower than the homicide arrest rate in 2014. 
```{r}
72/320*100
```

##Statement 3: Of 50 of the nation’s largest cities, Baltimore is one of 34 where police now make homicide arrests less often than in 2014, according to a Washington Post analysis.

Step 1: Create data frame that calculates homicides per city in 2014
```{r} 
homicide_city_2014 <- homicide_data |> 
  filter(reported_year == '2014') |>
  group_by(city) |> 
  summarize(count_homicides= n()) |> 
  arrange(desc(count_homicides))

homicide_city_2014
```
Step 2: Creating data frame that has homicides arrest per city in 2014
```{r} 
arrest_2014 <- homicide_data |> 
  filter(reported_year == '2014') |>
  filter(disposition == "Closed by arrest") |>
  group_by(city) |> 
  summarize(cases_closed = n()) |> 
  arrange(desc(cases_closed))
 
arrest_2014
```

Step 3: Creating data fram with homicides per city in 2017
```{r}
homicide_2017 <- homicide_data |>
  filter(reported_year == '2017') |>
  group_by(city) |>
  summarise(
    count_homicides =n()
  ) |>
  arrange(desc(count_homicides))
```

Step 4: Creating data frame that has closed by arrest per city in 2017

125 non arrests in 2014 out of 211 homicides, 59.24% no arrest

Data for 2017-

```{r}
arrest_2017 <- homicide_data |> 
  filter(reported_year == '2017') |>
  filter(disposition == "Closed by arrest") |>
  group_by(city) |>
  summarise(
    cases_closed=n()
  ) |>
  arrange(desc(cases_closed))
```

join the data sets from 2014 and 2017 then join the two into one big data frame
```{r}

```
```{r}

```
```{r}

```
```{r}

```


