---
title: "WP_Police_Homicides"
author: "Bridget Shaela Robert Sydney"
date: "2023-10-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

(Loading of the tidyverse)

```{r}
library(tidyverse)
library(lubridate)
library(janitor)
library(refinr)
```

(loading the data into R and correcting the dates, making a separate year column to assist when figuring out further questions.)
```{r}
homicide_data <- read_csv("data/homicide-data.csv") |>
mutate(reported_date=ymd(reported_date)) |>
  mutate(reported_year = year(reported_date))
```
## Statement 1: As Baltimore has seen a stunning surge of violence, with nearly a killing each day for the past three years in a city of 600,000, homicide arrests have plummeted. 

(Checking to see if this statement is true, if so there should be about 1,095 killings that happened within a three year period)

Step 1: filtering to have a data set with only Baltimore in it.)
```{r}
baltimore <- homicide_data |>
  filter(city == "Baltimore")
```

Step 2: Calculating # of homicides per year
```{r}
baltimore |>
 group_by(reported_year) |>
  summarise(
    count=n()
  ) |>
  arrange(reported_year)
```
Answer: Total deaths over three year span (2015, 2016, 2017) was 1,002. That's close to 1,095 which we predicted would be the range of killings. 

(Checking the number of homicides divided by the number of days in the past three years.)
```{r}
1002/365*3
```

##Statement 2: City police made an arrest in 41 percent of homicides in 2014; last year, the rate was just 27 percent, a 14 percentage point drop.

Step 1: Finding homicides in 2014
Answer: 211 Homicides in Baltimore in 2014
```{r}
baltimore |>
  filter(reported_year >= '2014') |>
  group_by(reported_year) |>
  summarise(
    count_homicides=n()
  )

```

Step 2: Finding number of arrests made in homicide cases in 2014
Answer: 86
```{r}
baltimore |>
  filter(reported_year >= '2014') |>
  filter(disposition == 'Closed by arrest') |>
  group_by(reported_year) |>
  summarise (
    count_arrests=n()
  )
  
```

Step 3: Divide total arrests by total homicides
```{r}
86/211*100
```
Answer: The Washington Post correctly calculated the arrest rate for homicides in the city of Baltimore in 2014 as 40.75 or roughly 41%. 


Now lets calculate the arrest rate for homicides in 2017. Was it really 27% as mentioned in the article? 


Step 1: Diving total arrests by total homicides for 2017. Referencing the previous data.
Answer: It's correct the arrest rate was 27.35 or roughly 27%. 

```{r}
93/340*100
```

Now we want to see if homicide arrest rates plummeted for 2015 and 2016. 

Step 1: Calculate 2015 using data from previous coding. 
Answer: 25.4%
```{r}
87/342*100
```

Step 2: Calculating 2016
Answer: 23% which is lower than the homicide arrest rate in 2014. 
```{r}
72/320*100
```

##Statement 3: Of 50 of the nation’s largest cities, Baltimore is one of 34 where police now make homicide arrests less often than in 2014, according to a Washington Post analysis.

Step 1: Create data frame that calculates homicides per city in 2014
```{r} 
homicide_city_2014 <- homicide_data |> 
  filter(reported_year == '2014') |>
  group_by(city) |> 
  summarize(count_homicides= n()) |> 
  arrange(desc(count_homicides))

homicide_city_2014
```

Step 2: Creating data frame that has homicide arrests per city in 2014
```{r} 
arrest_2014 <- homicide_data |> 
  filter(reported_year == '2014') |>
  filter(disposition == "Closed by arrest") |>
  group_by(city) |> 
  summarize(cases_closed = n()) |> 
  arrange(desc(cases_closed))
 
arrest_2014
```

Step 3: Creating data frame with homicides per city in 2017
```{r}
homicide_2017 <- homicide_data |>
  filter(reported_year == '2017') |>
  group_by(city) |>
  summarise(
    count_homicides =n()
  ) |>
  arrange(desc(count_homicides))
```

Step 4: Creating data frame that has homicide arrests per city in 2017
```{r}
arrest_2017 <- homicide_data |> 
  filter(reported_year == '2017') |>
  filter(disposition == "Closed by arrest") |>
  group_by(city) |>
  summarise(
    cases_closed=n()
  ) |>
  arrange(desc(cases_closed))
```

Step 5: Joining the 2014 data frames
```{r}
arrest_rate_2014 <- homicide_city_2014 |> left_join(arrest_2014, join_by(city)) |>
  mutate(arrest_rate = (cases_closed/count_homicides)*100)

arrest_rate_2014
```

Step 6: Joining the 2017 data frames
```{r}
arrest_rate_2017 <- homicide_2017 |> left_join(arrest_2017, join_by(city)) |>
  mutate(arrest_rate = (cases_closed/count_homicides)*100)
```

Step 7: Combine the 2014 + 2017 data to compare the arrest rates and see if there are about 34 cities who's rate have decreased. 
```{r}
all_arrest_rate <- arrest_rate_2014 |> left_join(arrest_rate_2017, join_by(city)) |>
rename(total_homicides_2014 = count_homicides.x, closed_cases_2014 = cases_closed.x, arrest_rate_2014 = arrest_rate.x, total_homicides_2017 = count_homicides.y, closed_cases_2017 = cases_closed.y, arrest_rate_2017 = arrest_rate.y)

```
Answer: We can confirm that there are 34 cities, including Baltimore who's arrest rate decreased. 

## Statement 4: And while most cities saw their arrest rates drop gradually, Baltimore’s decline was sudden — plummeting 15 percentage points in 2015, after Gray’s death, the largest single-year drop for any city already solving less than half its homicides.

Step 1: Create data frame for all homicides per city in 2015
```{r}
city_homicide_2015 <- homicide_data |>
  filter(reported_year == '2015') |>
  group_by(city) |>
  summarise (homicides = n()) |>
  arrange(desc(homicides))
```

Step 2: Create data frame for all homicide cases closed by arrest in 2015.
```{r}
city_closed_2015 <- homicide_data |>
  filter(reported_year == '2015') |>
  filter(disposition == "Closed by arrest") |>
  group_by(city) |>
  summarise(cases_closed=n()) |>
  arrange(desc(cases_closed))
```

Step 3: Join the data frames and create a column called "arrest rate" to display the arrest rate in 2015 for all cities
```{r}
arrest_rate_2015 <- city_homicide_2015 |> left_join(city_closed_2015, join_by(city)) |>
  mutate(arrest_rate = (cases_closed/homicides)*100) 
```

Step 4: Now let's compare them to previous years
```{r}
arrest_rate_change <- arrest_rate_2014 |> left_join(arrest_rate_2015, join_by(city)) |>
  rename(total_homicides_2014 = count_homicides, closed_cases_2014 = cases_closed.x, arrest_rate_2014 = arrest_rate.x, total_homicides_2017 = homicides, closed_cases_2017 = cases_closed.y, arrest_rate_2017 = arrest_rate.y)
```

Step 5: Subtract arrest rate of 2014 to 2015
```{r}
40-25
```
Answer: We can prove that arrest rates in Baltimore dropped by 15 percentage points following the death of Freddie Gray.


##Statement 5: For most of the decade before 2015, Baltimore’s annual homicide arrest rate hovered at about 40 percent. 

Step 1: Create new data frame to show number of homicides in Baltimore from 2007-2014. We already have a data frame with the Baltimore homicide data, so now we're filtering to look at a specific time period. 
```{r}
seven_year_baltimore_homicide <- baltimore |>
  filter(reported_year < '2015') |>
  group_by(reported_year) |>
  summarise (count_homicides = n())
```
Step 2: Create data frame for seven years of homicide arrests
```{r}
seven_year_baltimore_arrest <- baltimore |>
  filter(reported_year < '2015') |>
  filter(disposition == "Closed by arrest") |>
  group_by(reported_year) |>
  summarise(count_arrest=n())
```
Step 3: Join the two data frames and create a column with the arrest rate
```{r}
seven_year_baltimore_homicide_arrests <- seven_year_baltimore_homicide |> left_join(seven_year_baltimore_arrest, join_by(reported_year)) |>
  mutate(seven_year_arrest_rate = (count_arrest/count_homicides)*100)
```
Answer: After completing the data, we can confirm that the average arrest rate from 2007-2014 hovered around 40%.
