---
title: "lab_03"
author: "Shaela Foster"
date: "2023-09-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

library(tidyverse)
# Load the tidyverse
library(janitor)
# Load janitor
```

## TASK: Load data and clean up the column names

Load the two datasets we worked with in the pre_lab: the Maryland payments to vendors and Maryland grants and loans and, using janitor, change the columns so that they are lower-cased and any spaces are replaced with underscores.

```{r}
# Load in the Maryland state agency payments data, making sure that it correctly sets the datatypes using all of the rows
payments <- read_csv("data/State_of_Maryland_Payments_Data__FY2008_to_FY2024.csv", guess_max=373564)
md_grants_loans <- read_csv("data/State_of_Maryland_Grant_and_Loan_Data__FY2009_to_FY2022.csv")

cleaned_md_grants_loans <- md_grants_loans |>
  clean_names()

payments_cleaned <- payments |>
  clean_names()
# Load in the Maryland grants and loan data
```

## Answer questions

Q1: You want to know how much money state agencies have paid to the Prince George's County School system, but to do that you'll need to find all of the variations of how the school system might be listed. 
1. Write code to find rows where the vendor name contains the county's school system, starting with a broad search and then narrowing it down. 
2. When you have found as many as possible, save the result to a new dataframe called `pg_schools`. 
3. How many records did you find? Look at & search the records in the dataframe while you construct your code.

A1: I found records from 2008-2022.

```{r}
pg_schools <- payments_cleaned |>
  filter(
    str_detect(vendor_name, "BOARD OF EDUCATION OF PRINCE GEORGE'S COUNTY") | 
    str_detect(vendor_name, "PG COUNTY PUBLIC SCHOOL") |
    str_detect(vendor_name, "BD OF EDUCATION OF PG CO") |
    str_detect(vendor_name, "PRINCE GEORGES COUNTY PUBLIC SCHOOLS") |
    str_detect(vendor_name, "PRINCE GEORGES COUNTY PUBLIC SCHOOL") |
    str_detect(vendor_name, "BOARD OF EDUCATION OF PRINCE GEORGES COUNTY")
    )



pg_schools

```

Q2: Using that `pg_schools` dataframe, write code to calculate the total amount of payments to the Prince George's County School system for each fiscal year. Which fiscal year had the largest total amount? Are there any aspects of the results that stand out to you?

A2: In 2022, PG County Public Schools received the most payments. Something that stands out is that in 2016, they received the third most payments, I wonder why that is? I would assume that from every year 2008 to now it should steadily be increasing but that isn't the case with this data. 

```{r}
pg_schools |>
  group_by(fiscal_year) |>
  summarise(
    total_amount = sum(amount)
  ) |>
  arrange(desc(total_amount))

pg_schools

```

Q3. You're interested in examining the vendors paid for capital projects at the University of Maryland. 
1. Search the payments data to find those payments, 
2. then write code to produce a list of each vendor and the total amount it received. 
3. Look carefully at that list - what cleaning will it require to ensure that you have only one row per vendor and zip code? Provide at least three examples of vendor names that are similar and should be combined into one. Describe how you might clean that data so that any nearly identical vendors would look identical - what would you change and how?

A3. Three vendor names that are similar and should be combined are A MORTON THOMAS & ASSOC INC and A MORTON THOMAS & ASSOCATIONS INC, AMERICAN DESIGN ASSOCIATE and AMERICAN DESIGN ASSOCIATES, and AMTEK COMPANY, INC. and AMTEK COMPANY INC . I might clean this data up by using "case when" to make the companies with similar names under one name. Then I would add the amounts up from each company to find a grand total. From there, I could probably delete the other rows. Also, I could probably do the same case when with the zip codes so they match up.
```{r}
UMD_projects <- payments_cleaned |>
  filter (
    str_detect (agency_name, "CAPITAL PROJECTS- UNIVERSITY OF MARYLAND")) |>
group_by(vendor_name, vendor_zip) |>
summarise(
    total_amount = sum(amount)
  )
  


UMD_projects
```

Q4: You want to understand how much state grant money went to stem cell research projects each fiscal year.
1. Search the data to find grants that fit that description. 
2. Once you find which columns contain the information you want, your code should standardize the data needed to find just the grants for stem cell research in a single filter. 
3. Then create a total amount of money for each fiscal year. 
4. Write a sentence summarizing how stem cell research funding has changed over time.

A4: Based on the data, stem cell research funding has decreased over time. In 2009, it saw the most funding but it went down drastically the next year. Around when the pandemic hit, there was a slight increase in funding then it went back down in 2021. Would this be because of the lack of workers or money being generated throughout the lockdown?

```{r}
stem_cell <- cleaned_md_grants_loans |>
  filter(
    str_detect(description, "Stem Cell Research Grant") |
    str_detect(description, "STEM CELL RESEARCH GRANT") |
    str_detect(description, "Stem Cell Research Grant, Pre-Commercialization Research")
    ) |>
  group_by(fiscal_year) |>
      summarise(
        total_amount = sum(amount))
  

stem_cell
``` 

Q5: Using the Maryland grants and loan data, 
1. find records where the zip code is the one where the University of Maryland-College Park is located in and
2. produce a list of grantees from that zip code, along with the total amount they received. 
3. Write a paragraph explaining how many grantees are there, what you would need to do to clean the results so that you had one record per grantee and which grantees are most interesting to you?

A5: There are 17 grantees that have given money in the UMD-College Park area. Majority of the names are very similar and they have the University of Maryland as the name, but some list more information, spell them differently or format them differently. I would clean up the data by string detecting to have all the names in lowercase then somehow using "case when" to tell the data to recognize all the various version of UMD and place them under one name. Hopefully from this, it would make it easier to see which ones from directly from UMD. The one most interesting is the GripBoost. I looked it up and it's a glove making company that creates athletic gloves for sports like football and ice hockey. They granted about $100,000 which was very interesting because I expected more educational related grants for research or something similar, not anything related to athletics. 

```{r}
UMD_zipcode <- cleaned_md_grants_loans |>
  filter(zip_code == 20742) |>
  group_by(grantee) |>
  summarise(
    total_amount = sum(amount)
  ) |>
  arrange(desc(total_amount))

UMD_zipcode
```

