---
title: "lab_05"
author: "Shaela Foster"
date: "2023-10-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   Tabula

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

library(tidyverse)
library(lubridate)
library(janitor)
# Load the tidyverse, plus any other packages you will need to clean data and work with dates.

```

## Get Our PDF

We'll be working with the [911 overdose calls from Baltimore County](https://drive.google.com/file/d/1qkYuojGF_6WKFr5aNQxmewDzcKyOiJFr/view?usp=share_link). You'll want to download it to a place you'll remember (like your Downloads folder, or the labs folder in your repository). The goal is to extract the tables within it, export that to a CSV file, load it into RStudio and ask some questions.

## Extract Data from PDF Using Tabula

Start Tabula, then go to <http://127.0.0.1:8080/> in your browser. Click the "Browse" button and find the PDF file and click "open", and then click the "Import button" in Tabula. This will take a few seconds or longer.

This PDF has a single table spread over multiple pages to extract. We're going to make a single dataframe from this table, exporting it to a CSV file that you will load into R. In Tabula, highlight the table and click the "Preview & Export Extracted Data" button. You may want to play with including or excluding the column headers - YOU SHOULD HAVE FIVE COLUMNS OF DATA.

Save the CSV (it should be called `tabula-Baltimore County; Carey, Samantha log OD.csv` by default) to your lab_05/data folder.

From there, you will need to read in the data, and add or fix headers if necessary. You can choose to include the headers from the PDF in your exported CSV files OR to exclude them and add them when importing. `read_csv` allows us to do this ([and more](https://readr.tidyverse.org/reference/read_delim.html)).

## Load and clean up the data in R

You will need to read in and clean up the data so that it can be used for analysis. By "clean" I mean the column headers should not contain spaces and they should have meaningful names, not "x1" or something similar. How you do that is up to you, but you can use select() with or without the minus sign to include or exclude certain columns. You also can use the `rename` function to, well, rename columns. Importantly, you'll need to ensure that any columns containing a date actually have a date datatype. Our friend `lubridate` can help with this.

```{r}
overdose_data <- read_csv("data/tabula-Baltimore County; Carey, Samantha log OD.csv", col_names = FALSE) |> clean_names() |>
  rename(date = x1, time = x2, case_number = x3, EVTYP = x4, location = x5) |>
  mutate(date=mdy(date))
```

## Answer questions

Q1. Write code to generate the number of calls that occurred on each date. Which date in 2022 had the most overdose calls, and how many? Look at the total number of rows in your result and explore the range of dates - based on your result, do you believe there are any days with no overdose calls at all? Explain why or why not.

A1. July 14 and October 4 had the most overdose calls in 2022 with 23 each. Based on the data, there are no overdose calls for the entire month of January in 2022. Not sure if no one called during this month or if they didn't start recording data until February 6. 

```{r}
overdose_data |>
  filter(date < "2023-01-01") |>
   group_by(date) |>
  summarise(
    count = n()
  ) |>
  arrange(desc(count))
```

Q2. You want to understand if there's a pattern in the day of the week that overdose calls are made. Add a column to your dataframe that displays what day of the week each date represents. You should search for how to do that using lubridate. Then write code to calculate the number of calls for each day of the week, and add a column to that result that calculates the percentage of all calls that occurred on each day of the week 
1. so you want a dataframe with the day of the week, 
2. total number of calls and the 
3. (percentage of calls on that day out of the total number of all calls). Describe your findings to me.

A2. I found that Saturday's are the days where the most overdose calls occur making about 16% of the average calls. Thursday's are the days with the least number of calls with 526 calls making up about 13% of calls. 

```{r}
days <- overdose_data |>
  mutate(weekday = wday(date, label = TRUE, abbr = FALSE)) |>
group_by(weekday) |>
  summarise (
    count=n(),
    count/4112*100
  )
  
days
```

Q3. Now let's look at locations. Which ones have the most calls? How would you describe them (feel free to search for more information on them)? Is there anything about the structure of the original data that might make you less confident in the counts by location or date?

A3. On the original data set, we're missing dates from January 2022, the locations don't have full addresses so maybe providing the zip codes could help group some of these things so we could see what areas has the most calls. Although I can look the addresses up, they show the what the homes look like now and not what it looked like previously so for me it causes some skepticism on if its the correct location. 
```{r}
locations <- overdose_data |>
  group_by(location) |>
  summarise(
    count=n()
  ) |>
  arrange(desc(count))
locations
```

Q4. What's the best story idea or question you've seen as a result of the work you've done in this lab?

A4. Possible story ideas would be looking at the homes that have multiple overdose calls, have been renovated, and how much they are on the market for if they are still attempting to be sold. Another story could be looking at the overdose calls that occur at the precinct and looking at when the overdoses are happening, like while people are in custody or anything surrounding that process. Also, if there are officers involved in these cases. I found a case from Woodlawn HS, so a story could be analyzing the programs high schools have put in place preventing students from overdosing or using drugs of any kind. 
